---
title: "R Notebook"
output: html_notebook
---

The goal of this project is to investigate how partnerships involving multiple top-tier players in the NBA impacts various performance measures and team outcomes. Among the research questions we would like to explore are the following:

*[ INSERT RESEARCH QUESTIONS HERE ]*

To be able to investigate, we need to pull data from multiple NBA seasons. The script below provides code to create functions that pull traditional stats for every player for a given user-defined season.

Load necessary libraries
```{r}
library(rvest)
library(dplyr)
library(tidyverse)
library(httr)

```

Function to get NBA roster for a specified year


Example usage





**ASSIGNMENT 1:** *Is the data "clean"? Are there any missing values to be accounted for/addressed? If there are any data quality issues,*

 - *a. propose a method to resolve them*
       
Initial thought to change the default character to double given that we have fractioned values. i think the columns should be changed from " chararcter" to "double" 


 - *b. justify the validity of your approach*
removing observations with missing data from the dataset, using the function "na.omit" which will remove rows with missing values from our dataset


 - *c. implement your proposed changes*


**ASSIGNMENT 2:** *Is the advanced data "clean"? Are there any missing values to be accounted for/addressed? If there are any data quality issues,*

 - *a. propose a method to resolve them*

 - *b. justify the validity of your approach*

 - *c. implement your proposed changes*
 
 cleaning similar to first one 
 
 
 The script below provide code to clean out the quality issues presented in the dataframe
 
 
 
```{r}

```
 
 
 
```{r}

```


**ASSIGNMENT 3:** *Merge the cleaned up datasets to create one new data frame with the traditional and advanced stats.*



```{r}
#Get NBA Totals Statistics
get_nba_totals_stats <- function(year) {
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", year, "_totals.html")
  webpage <- read_html(url)
  totals_stats_table <- webpage %>%
    html_node("table#totals_stats") %>%
    html_table(fill = TRUE)

  # Clean up column names
  colnames(totals_stats_table) <- make.names(colnames(totals_stats_table), unique = TRUE)

  # Clean the data
  totals_stats_table <- totals_stats_table %>%
    filter(!is.na(Player) & Player != "Player")  # Ensure no NA or duplicate header rows
  
  return(totals_stats_table)
}


get_nba_advanced_stats <- function(year) {
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", year, "_advanced.html")
  
  # Fetch webpage
  webpage <- tryCatch({
    read_html(GET(url, user_agent("Mozilla/5.0")))
  }, error = function(e) {
    stop("Error fetching webpage: ", e$message)
  })
  
  # Extract table with updated ID
  advanced_stats_table <- webpage %>%
    html_node("table#advanced") %>%  # Updated selector to match the new ID
    html_table(fill = TRUE)
  
  # Clean up column names
  colnames(advanced_stats_table) <- make.names(colnames(advanced_stats_table), unique = TRUE)
  
  # Clean the data
  advanced_stats_table <- advanced_stats_table %>%
    filter(!is.na(Player) & Player != "Player")  # Remove NA rows and duplicate headers
  
  return(advanced_stats_table)
}


```


**ASSIGNMENT 4:** *Make a function with argument `year` that outputs one dataframe with the merged traditional and advanced data.* 


Official Cleaning Function that works as of 10/29/2024
```{r}

#Get NBA Totals Statistics
get_nba_totals_stats <- function(year) {
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", year, "_totals.html")
  webpage <- read_html(url)
  totals_stats_table <- webpage %>%
    html_node("table#totals_stats") %>%
    html_table(fill = TRUE)

  # Clean up column names
  colnames(totals_stats_table) <- make.names(colnames(totals_stats_table), unique = TRUE)

  # Clean the data
  totals_stats_table <- totals_stats_table %>%
    filter(!is.na(Player) & Player != "Player")  # Ensure no NA or duplicate header rows
  
  return(totals_stats_table)
}


get_nba_advanced_stats <- function(year) {
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", year, "_advanced.html")
  
  # Fetch webpage
  webpage <- tryCatch({
    read_html(GET(url, user_agent("Mozilla/5.0")))
  }, error = function(e) {
    stop("Error fetching webpage: ", e$message)
  })
  
  # Extract table with updated ID
  advanced_stats_table <- webpage %>%
    html_node("table#advanced") %>%  # Updated selector to match the new ID
    html_table(fill = TRUE)
  
  # Clean up column names
  colnames(advanced_stats_table) <- make.names(colnames(advanced_stats_table), unique = TRUE)
  
  # Clean the data
  advanced_stats_table <- advanced_stats_table %>%
    filter(!is.na(Player) & Player != "Player")  # Remove NA rows and duplicate headers
  
  return(advanced_stats_table)
}

get_cleaned_nba_stats <- function(year) {
  # Fetch totals and advanced stats
  nba_totals <- get_nba_totals_stats(year)
  nba_advanced <- get_nba_advanced_stats(year)

  # Print to check datasets (Optional)
  print("NBA Totals:")
  print(head(nba_totals))
  print("NBA Advanced:")
  print(head(nba_advanced))

  # Clean Player names in the advanced dataset: remove the asterisk and trim spaces
  nba_advanced <- nba_advanced %>%
    mutate(Player = trimws(gsub("\\*", "", Player)))  # Remove asterisk

  # Ensure that the advanced stats consider the cleaned player names
  nba_advanced <- nba_advanced %>%
    mutate(Player = trimws(Player))

  # Merge the datasets on Player, Pos, G, and MP
  nba_merge <- merge(nba_totals, nba_advanced, 
                     by = c("Player", "Pos", "G", "MP"), 
                     all.x = TRUE)

  # Debug: Show merged data sample
  print("Merged Data Sample:")
  print(head(nba_merge))
  
  # Check column names to confirm 'Team' exists
  print("Column Names Before Renaming:")
  print(colnames(nba_merge))

  # Clean 'Team' column: rename 'Tm' to 'Team' if present
  if ("Tm" %in% colnames(nba_merge)) {
    nba_merge <- nba_merge %>%
      rename(Team = Tm)
  }

  # Debug: Show column names after renaming
  print("Column Names After Renaming 'Tm' to 'Team':")
  print(colnames(nba_merge))

  # Handle duplicate columns like Team.x, Team.y, Awards.x, Awards.y
  duplicate_columns <- colnames(nba_merge)[grepl("\\.x$", colnames(nba_merge))]

  for (col in duplicate_columns) {
    # Extract the base name of the column (e.g., "Team" from "Team.x")
    base_col <- sub("\\.x$", "", col)
    
    # Merge the .x and .y columns into one
    if (paste0(base_col, ".y") %in% colnames(nba_merge)) {
      nba_merge <- nba_merge %>%
        mutate(!!base_col := coalesce(get(col), get(paste0(base_col, ".y")))) %>%
        select(-all_of(c(col, paste0(base_col, ".y"))))  # Drop the old columns
    }
  }

  # Remove players whose team is "TOT", "2Tm", or "3Tm"
  nba_merge <- nba_merge %>%
    filter(!grepl("^(TOT|2TM|3TM)$", Team))

  # Remove players with multiple positions
  nba_merge <- nba_merge %>%
    filter(!grepl("-", Pos))

  # Remove the 'X', 'X.1' columns if they exist
  columns_to_remove <- c("X", "X.1")
  nba_merge <- nba_merge %>%
    select(-any_of(columns_to_remove))  # Remove specified columns if they exist

  # Merge 'Rk.x' and 'Rk.y' columns
  if ("Rk.x" %in% names(nba_merge) & "Rk.y" %in% names(nba_merge)) {
    nba_merge <- nba_merge %>%
      mutate(Rk = coalesce(as.character(Rk.x), as.character(Rk.y))) %>%
      select(-Rk.x, -Rk.y)
  }

  # Merge 'Age.x' and 'Age.y' columns
  if ("Age.x" %in% names(nba_merge) & "Age.y" %in% names(nba_merge)) {
    nba_merge <- nba_merge %>%
      mutate(Age = coalesce(as.numeric(Age.x), as.numeric(Age.y))) %>%
      select(-c(Age.x, Age.y))
  }

  # Reorder columns for clarity
  column_order <- c("Player", "Pos", "Age", "Rk", "G", "MP", "Team")
  nba_merge <- nba_merge %>%
    select(all_of(column_order), everything())

  # Return the cleaned and merged dataset
  return(nba_merge)
}

# Example usage
nba_data_2013 <- get_cleaned_nba_stats(2013)

# View the first few rows of the cleaned dataset
head(nba_data_2013)


```


```{r}
nba_data_2022 <-get_cleaned_nba_stats(2022)
nba_data_2010 <-get_cleaned_nba_stats(2010)
nba_data_2015 <-get_cleaned_nba_stats(2015)
nba_data_2011<-get_cleaned_nba_stats(2011)
nba_data_2012<-get_cleaned_nba_stats(2012)
nba_data_2009<-get_cleaned_nba_stats(2009)
nba_data_2008<-get_cleaned_nba_stats(2008)
nba_data_2007<-get_cleaned_nba_stats(2007)
nba_data_2006<-get_cleaned_nba_stats(2006)
nba_data_2005<-get_cleaned_nba_stats(2005)
nba_data_2004<-get_cleaned_nba_stats(2004)
nba_data_2003<-get_cleaned_nba_stats(2003)
nba_data_2002<-get_cleaned_nba_stats(2002)
nba_data_2001<-get_cleaned_nba_stats(2001)
nba_data_2000<-get_cleaned_nba_stats(2000)



nba_data_1999<-get_cleaned_nba_stats(1999)
nba_data_1998<-get_cleaned_nba_stats(1998)
nba_data_1997<-get_cleaned_nba_stats(1997)
nba_data_1996<-get_cleaned_nba_stats(1996)
nba_data_1995<-get_cleaned_nba_stats(1995)
nba_data_1994<-get_cleaned_nba_stats(1994)
nba_data_1993<-get_cleaned_nba_stats(1993)
nba_data_1992<-get_cleaned_nba_stats(1992)
nba_data_1991<-get_cleaned_nba_stats(1991)
nba_data_1990<-get_cleaned_nba_stats(1990)
nba_data_1989<-get_cleaned_nba_stats(1989)
nba_data_1988<-get_cleaned_nba_stats(1988)
nba_data_1987<-get_cleaned_nba_stats(1987)
nba_data_1986<-get_cleaned_nba_stats(1986)
nba_data_1985<-get_cleaned_nba_stats(1985)
nba_data_1984<-get_cleaned_nba_stats(1984)
nba_data_1983<-get_cleaned_nba_stats(1983)
nba_data_1982<-get_cleaned_nba_stats(1982)
nba_data_1981<-get_cleaned_nba_stats(1981)
nba_data_1980<-get_cleaned_nba_stats(1980)

readLines("https://www.basketball-reference.com/teams/LAL/2023.html", n = 1)

```


**ASSIGNMENT 5:** *Make this file more visually appealng, with headers, bullet points, sections and subsections as you see fit. You may consider migrating over to Quarto for this reason.*


File locator
```{r}
# Save your dataframe as a CSV file
write.csv(nba_roster2, file = "generalstats.csv", row.names = FALSE)
write.csv(AO_nba_advanced_stats2, file = "advancedstats.csv", row.names = FALSE)
write.csv(nba_data_2023, file = "nba2023.csv", row.names = FALSE)
write.csv(nba_data_2013, file = "nba2013.csv", row.names = FALSE)
getwd()

```





