---
title: "R Notebook"
output: html_notebook
---

The goal of this project is to investigate how partnerships involving multiple top-tier players in the NBA impacts various performance measures and team outcomes. Among the research questions we would like to explore are the following:

*[ INSERT RESEARCH QUESTIONS HERE ]*

To be able to investigate, we need to pull data from multiple NBA seasons. The script below provides code to create functions that pull traditional stats for every player for a given user-defined season.

Load necessary libraries
```{r}
library(rvest)
library(dplyr)
library(tidyverse)
```

Function to get NBA roster for a specified year


Example usage





**ASSIGNMENT 1:** *Is the data "clean"? Are there any missing values to be accounted for/addressed? If there are any data quality issues,*

 - *a. propose a method to resolve them*
       
Initial thought to change the default character to double given that we have fractioned values. i think the columns should be changed from " chararcter" to "double" 


 - *b. justify the validity of your approach*
removing observations with missing data from the dataset, using the function "na.omit" which will remove rows with missing values from our dataset


 - *c. implement your proposed changes*


**ASSIGNMENT 2:** *Is the advanced data "clean"? Are there any missing values to be accounted for/addressed? If there are any data quality issues,*

 - *a. propose a method to resolve them*

 - *b. justify the validity of your approach*

 - *c. implement your proposed changes*
 
 cleaning similar to first one 
 
 
 The script below provide code to clean out the quality issues presented in the dataframe
 
 
 
```{r}

```
 
 
 
```{r}



```


**ASSIGNMENT 3:** *Merge the cleaned up datasets to create one new data frame with the traditional and advanced stats.*



```{r}
#Get NBA Totals Statistics
get_nba_totals_stats <- function(year) {
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", year, "_totals.html")
  webpage <- read_html(url)
  totals_stats_table <- webpage %>%
    html_node("table#totals_stats") %>%
    html_table(fill = TRUE)

  # Clean up column names
  colnames(totals_stats_table) <- make.names(colnames(totals_stats_table), unique = TRUE)

  # Clean the data
  totals_stats_table <- totals_stats_table %>%
    filter(!is.na(Player) & Player != "Player")  # Ensure no NA or duplicate header rows
  
  return(totals_stats_table)
}


#get NBA advanced stats
get_nba_advanced_stats <- function(year) {
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", year, "_advanced.html")
  webpage <- read_html(url)
  advanced_stats_table <- webpage %>%
    html_node("table#advanced_stats") %>%
    html_table(fill = TRUE)
  
  # Clean up column names
  colnames(advanced_stats_table) <- make.names(colnames(advanced_stats_table), unique = TRUE)
  
  # Clean the data
  advanced_stats_table <- advanced_stats_table %>%
    filter(!is.na(Player) & Player != "Player") # Ensure no NA or duplicate header rows
  
  return(advanced_stats_table)
}

```


**ASSIGNMENT 4:** *Make a function with argument `year` that outputs one dataframe with the merged traditional and advanced data.* 


Official Cleaning Function that works as of 10/29/2024
```{r}
# Function to merge NBA totals and advanced stats for a specified year
get_cleaned_nba_stats <- function(year) {
  # Fetch totals and advanced stats
  nba_totals <- get_nba_totals_stats(year)
  
  nba_advanced <- get_nba_advanced_stats(year)

  # Print to check datasets
  print("NBA Totals:")
  print(head(nba_totals))
  print("NBA Advanced:")
  print(head(nba_advanced))

  # Clean Player names in the advanced dataset: remove the asterisk and trim spaces
  nba_advanced <- nba_advanced %>%
    mutate(Player = trimws(gsub("\\*", "", Player)))  # Remove asterisk

  # Ensure that the advanced stats consider the cleaned player names
  nba_advanced <- nba_advanced %>%
    mutate(Player = trimws(Player))

  # Merge the datasets on Player, Pos, G, and MP
  nba_merge <- merge(nba_totals, nba_advanced, 
                     by = c("Player", "Pos", "G", "MP"), 
                     all.x = TRUE)

  # Debug: Show merged data sample
  print("Merged Data Sample:")
  print(head(nba_merge))

  # Remove players whose team is "TOT", "2Tm", or "3Tm"
  nba_merge <- nba_merge %>%
    filter(!grepl("^(TOT|2TM|3TM)$", Team))

  # Remove players with multiple positions
  nba_merge <- nba_merge %>%
    filter(!grepl("-", Pos))

  # Remove the 'Tm', 'X', and 'X.1' columns if they exist
  columns_to_remove <- c("Tm", "X", "X.1")
  nba_merge <- nba_merge %>%
    select(-any_of(columns_to_remove))  # Remove specified columns if they exist

  # Merge Rk.x and Rk.y
  if ("Rk.x" %in% names(nba_merge) & "Rk.y" %in% names(nba_merge)) {
    nba_merge <- nba_merge %>%
      mutate(Rk = coalesce(as.character(Rk.x), as.character(Rk.y))) %>%
      select(-Rk.x, -Rk.y)
  }

  # Merge Age columns
  if ("Age.x" %in% names(nba_merge) & "Age.y" %in% names(nba_merge)) {
    nba_merge <- nba_merge %>%
      mutate(Age = coalesce(as.numeric(Age.x), as.numeric(Age.y))) %>%
      select(-c(Age.x, Age.y))
  }

  # Reorder columns for clarity
  column_order <- c("Player", "Pos", "Age", "Rk", "G", "MP", "Team")
  nba_merge <- nba_merge %>%
    select(all_of(column_order), everything())

  return(nba_merge)
}

# Example usage
nba_data_2013 <- get_cleaned_nba_stats(2013)

# View the first few rows of the cleaned dataset
head(nba_data_2013)


```


```{r}
nba_data_2022 <-get_cleaned_nba_stats(2022)
nba_data_2010 <-get_cleaned_nba_stats(2010)
nba_data_2015 <-get_cleaned_nba_stats(2015)
```


**ASSIGNMENT 5:** *Make this file more visually appealng, with headers, bullet points, sections and subsections as you see fit. You may consider migrating over to Quarto for this reason.*


File locator
```{r}
# Save your dataframe as a CSV file
write.csv(nba_roster2, file = "generalstats.csv", row.names = FALSE)
write.csv(AO_nba_advanced_stats2, file = "advancedstats.csv", row.names = FALSE)
write.csv(nba_data_2023, file = "nba2023.csv", row.names = FALSE)
write.csv(nba_data_2013, file = "nba2013.csv", row.names = FALSE)
getwd()

```


